---
title: 'Indianapolis: Crime Downtown'
subtitle: 'Is it dangerous?'
date: '2024-01-13'
categories: [post]
execute: 
  warning: false
format:
  html:
    toc: true
---

Indianapolis is smack dab in the middle of Indiana, which is a conservative state. Therefore, conservative talk radio is abundant. One of my guilty pleasures is conservative talk radio. I can't count the number of times my father had Rush Limbaugh on in the car. The radioman was an additional, parasocial crazy uncle. I have always had some form of conservative media in my life. Hearing old men gasp at the latest liberal cultural transgression makes me feel nostologic & at home.

If you were to take these guys^[[Tony Kats](https://twitter.com/tonykatz), [Robert Evans](https://twitter.com/r3evans), & [Tony Kinnett](https://twitter.com/TheTonus)] seriously, you'd think Indianapolis was a war zone. It is true that Indianapolis has a lot of [violence](https://wibc.com/153439/why-all-the-crime-in-indy-indy-ten-point-coalition-gives-his-thoughts/), so their warnings are not totally without merit. But when I hear them talk about the violence **in Indianapolis**, I think of downtown. 

I've lived in downtown Indianapolis for three years. I have never felt in danger. So, what gives? Maybe it's that I am a [large dude](https://ballstatesports.com/sports/football/roster/vinnie-palazeti/5071) (I've lost some weight since my playing days I swear). My wife does feel uncomfortable "near the circle". I've heard her mother, an Irvington resident, say similar things. 

Is downtown actually dangerous?

I am going to only consider Violent Crime, which in this dataset falls into one of these Categories: Homicide, Robbery, Assault, or Sexual Offense.

IMPD releases crime data with coordinates. This is the 2021 UCR data. Let's check it out. 


```{r}
#| echo: FALSE

# required libraries
library(tidycensus)
library(tidyverse)
library(sf)
options(tigris_use_cache = TRUE)
# census_api_key()
library(purrr)
library(leaflet)
library(kableExtra)
library(dplyr)
```

```{r}
#| echo: FALSE

# list of data files
directory <- paste(getwd(),'data',sep='/')
files<-list.files(directory, full.names=TRUE)
```

```{r}
#| echo: FALSE
# read in data, filter for Violent Crime and IMPD agency
# this data is from CityProtect
# it includes reports & police dispatches, not sure how to disaggregate that yet

# impd_files <- files[grepl('agency_109542',files)]
# impd_data <- do.call(rbind, lapply(impd_files, function(x) read.csv(x, stringsAsFactors = FALSE)))

# impd_data <- impd_data |>
#   distinct(id, .keep_all = TRUE) |> 
#   filter(parentIncidentType %in% c("Homicide","Robbery","Assault","Sexual Offense")) |>
#   drop_na(lat, lng) |> 
#   rename(LATITUDE=lat,LONGITUDE=lng,crime=incidentType) |>
#   filter(agencyId == "109542")


### UCR Data
impd_file <- files[grepl('IMPD_',files)]
impd_data <- read.csv(impd_file) |> mutate(CRIME = replace_na(CRIME, 'NA'))

# impd_data |>
#   count(CRIME) |> 
#   arrange(desc(n)) |>
#   head(30)

# impd_data |>
#   select(everything()) |>
#   summarise_all(funs(sum(is.na(.))))

violent_crime_str <- c("ASSAULT","ROBB","RAPE","HOMICIDE")
impd_data$violent_crime <- sapply(impd_data$CRIME, function(v) any(str_detect(v, violent_crime_str)))

impd_data <- impd_data |>
  filter(violent_crime == TRUE)

# impd_data |>
#   summarise(across(everything(), ~ sum(is.na(.))))

impd_data <- impd_data |> drop_na() 

# impd_data |>
#   count(CRIME) |> 
#   arrange(desc(n)) 

impd_data <- impd_data |> 
  mutate(
    crime = case_when(
      str_detect(CRIME,"ASSAULT") ~ "Assault",
      str_detect(CRIME,"HOMICIDE") ~ "Homicide",
      str_detect(CRIME,"ROBB") ~ "Robbery",
      str_detect(CRIME,"RAPE") ~ "Rape",
      TRUE ~ as.character(CRIME)
  ))
```


# Marion County

Here is a table of the violent crime for all of Marion County.

```{r}
impd_data |>
  count(CRIME) |> 
  arrange(desc(n)) |>
  rename(Occurances=n) |>
  kable()
```


Here is a cluster map of the individual crimes from Marion County. You can zoom in to find more info.

```{r}
#| echo: FALSE
marion <- get_acs(
  state = "IN",
  county = "Marion",
  geography = "tract",
  variables = "B01001_001",
  geometry = TRUE,
  year = 2022
) |> st_transform(crs = 4326)

marion$area <- st_area(marion)

points <- impd_data |> 
    select(crime, LONGITUDE, LATITUDE) |>
    sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326) |> 
    sf::st_transform(crs = 4326)

# marion |>
#   ggplot() + 
#   geom_sf() + 
#   geom_sf(data = points, color = alpha("red", 0.5)) + 
#   theme_minimal()

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircleMarkers(
    data=points,
    label = ~ crime,
    color='red',
    clusterOptions = markerClusterOptions())
```


# How to define downtown?

```{r}
#| echo: FALSE

# these are the tracts for downtown
donwtown_tracts <- c('Census Tract 3910.01','Census Tract 3910.02','Census Tract 3542.01','Census Tract 3542.02', 'Census Tract 3562')
format_tracts <- paste(donwtown_tracts, '; Marion County; Indiana', sep='')

# downtown tract tracts
downtown <- marion |>
    filter(NAME %in% format_tracts)

# not downtown tracts
not_downtown <- marion |>
    filter(!NAME %in% format_tracts) |> 
    filter(!is.na(estimate))

# each crime that is in downtown
downtown_crime <- points |> 
  st_join(downtown, left=FALSE) 

# each crime that is not in downtown
not_downtown_crime <- points |> 
  st_join(not_downtown, left=FALSE) 
```

When Hoosiers say downtown, I'd be willing to bet they mean somewhere close to the circle. I doubt most would consider Holy Cross, a neighbor just east of our boundary, downtown.

I am going to define downtown using US Census Tracts, which provide a nice boundary and population data. So, I will consider crimes that happen within this boundary as downtown crimes.

There were a total of `r nrow(downtown_crime)` violent crimes downtown last year.



```{r}
#| echo: FALSE
#| fig-cap: "Downtown Indianapolis: Census Tracts"
leaflet() |> 
  addTiles() |>
  addPolygons(data=downtown)
```



# What's the population of downtown?

According to the Census, the population of downtown is ~20,000 people. For our purposes, is this supposed to represent the number of people downtown all year? I have roughly a years worth of data, from January to January (2023 to 2024). Were there 20 thousand people in downtown Indianapolis throughout the year? That can't be right. On a Saturday during the summer I bet nearly 20,000 **extra** people are downtown.

According to Placer.ai, a website which estimates monthly visits to various locations, there are about ~300,000 monthly visits to zip code 46204, which contains much of downtown including the circle. Does that mean there were 300K x 12 (months) people downtown over the year?

Gainsbridge Fieldhouse hosts [2 million](https://www.gainbridgefieldhouse.com/plan-your-visit#:~:text=While%20the%20distinct%20exterior%20design,conventions%2C%20and%20games%20every%20year.) guests per year. How should we count this influx of people? 

Finally, the VisitIndy people say there are [30 million](https://www.visitindy.com/about-us/) annual visitors to Indianapolis. 

This is genuinely confusing, and I am not sure how to resolve it. If anyone has a good idea please let me know. 

The question I want to answer is: 

> How likely are you to be near or victim of a crime when visiting downtown? 
 
To answer this, I need the number of crimes per day divided by the number of people downtown per day. 

If I take VisitIndy's estimate at face value, that would mean on average there are 30,000,000 / 365 = ~82,000 people downtown. This seems way to high. The standard deviation, due to sports matches, holidays, and various events, is probably large. I'd be willing to bet the average is closer to 40,000 to 50,000 people who come to downtown each day. So, I'll split that range and call it 45,000. This will be my number of people in downtown Indianapolis each day.


```{r}
#| echo: FALSE
downtown_pop <- 45000
```


# How many crimes per day?


There are a total of 430 violent crimes downtown in the year. That is 1.17 violent crimes per day.

```{r}
#| echo: False
crime_per_day <- nrow(downtown_crime) / 365
```

# How many crimes per meter^2^?

The total area downtown is about 10 million meters^2^. A single square meter is about 10 square feet. There are 1.17 crimes downtown per day divided by 10 million meters^2^. That is equal to 0.0000001 crimes per day per meter^2^. 

Or, put another way, for any given ~3x3^[correction: initally read 10x10, credit to [josiah](https://twitter.com/@josiahkeller) for this correction] foot spot downtown you have a 1 in 10 million chance of being in the same spot as a violent crime. 

```{r}
downtown_area <- sum(downtown$area) 
# there are 0.000000119 crimes per day per meter downtown
# I like those odds
crimes_per_day_per_meter2 <- (crime_per_day / downtown_area)

crimes_per_day_per_meter2
```


# What is the crime per 100 people?

Back to population, there are ~0.002 violent crimes per day per 100 people downtown.

```{r}
# for every 100 people downtown, there are 0.002 violent crimes
(crime_per_day / downtown_pop)*100
```


# Is there more crime in the Circle?

My conservative uncles, and my beloved mother-in-law, may have the right intuition when it comes to the circle. The crime in the circle is higher than the rest of downtown.

Instead of **0.0000001 crimes** per day per meter^2^, there are **0.0000006 crimes** per day per meter^2^. The latter number is higher than the former. Is the difference significant? What do you think?

```{r}
#| echo: FALSE

# leaflet() |> 
#   addProviderTiles(providers$CartoDB.Positron) |> 
#   addCircleMarkers(
#     data=downtown_crime,
#     radius = 1,
#     fillOpacity = 0.5, 
#     color='red'
#     )

circle <- tibble(
    longitude = c(-86.1608, -86.15541, -86.15541, -86.1608),
    latitude = c(39.7705, 39.7705, 39.7665,39.7665)    
) 

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircleMarkers(
    label = ~ crime,
    data=downtown_crime,
    color='red',
    clusterOptions = markerClusterOptions()) |>
    addPolygons(lat=circle$latitude, lng=circle$longitude)


circle_poly <- circle |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_bbox() |>
  st_as_sfc() |>
  st_sf()

colnames(circle_poly) <- c('geometry')
st_geometry(circle_poly) <- 'geometry'
circle_poly$area <- st_area(circle_poly)

circle_crime <- downtown_crime |> st_join(circle_poly, left=FALSE)
circle_crime_per_day <- nrow(circle_crime) / 365
```

# What if you visted downtown every day?

If you lived for 80 years & visted downtown once every day, that would be roughly 29,200 visits. 

After your entire life, you would have had a 0.3% chance of being in the same spot as a violent crime

```{r}
years <- 80
total_visits <- years*365
odds <- (total_visits * crimes_per_day_per_meter2)
odds
```
